---
title: "ESM 206 Week 5 Coding Lab"
author: "Shelby Smith"
date: "11/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
library(effsize)
library(janitor)
library(broom)
library(ggbeeswarm)
library(car)
```

# Part 1. T-tests 

Mass comparison between male and female chinstrap penguins

```{r}
chinstrap_penguins <- penguins %>% 
  filter(species=="Chinstrap")

# Histograms (only takes one variable)
chinstrap_penguins %>% 
  ggplot(aes(x= body_mass_g)) +
  geom_histogram(bins=10)+
  facet_wrap(~sex)
  
```

- distribution overall looks pretty normal generally bell shaped, so mean would be a reasonable measure of central tendency

- male central tendency is higher (about 4000), and female central tendency is lower around 3500

## QQ plot

- #in qqplot, you tell r what the sample is aes(sample=)
```{r}
chinstrap_penguins %>% 
  ggplot(aes(sample=body_mass_g))+
  geom_qq()+
  facet_wrap(~sex)
```

- as close to normally distributed as possible 



## Get Descriptive Statistics

```{r}
descriptive_stats<- chinstrap_penguins %>% 
  group_by(sex) %>% 
  summarise(
    mean_mass= mean(body_mass_g),
    median_mass= median(body_mass_g),
    sd_mass= sd(body_mass_g),
    sample_size= n())
```

- means are a bit different for females and males as we saw in our histograms

- both have sample size of 34

- similar standard deviation

- median is similar to mean- refelcted in symmetric distribution in the histogram