---
title: "ESM 206 Week 5 Coding Lab"
author: "Shelby Smith"
date: "11/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
library(effsize)
library(janitor)
library(broom)
library(ggbeeswarm)
library(car)
```

# Part 1. T-tests 

Mass comparison between male and female chinstrap penguins

```{r}
chinstrap_penguins <- penguins %>% 
  filter(species=="Chinstrap")

# Histograms (only takes one variable)
chinstrap_penguins %>% 
  ggplot(aes(x= body_mass_g)) +
  geom_histogram(bins=10)+
  facet_wrap(~sex)
  
```

- distribution overall looks pretty normal generally bell shaped, so mean would be a reasonable measure of central tendency

- male central tendency is higher (about 4000), and female central tendency is lower around 3500

## QQ plot

- #in qqplot, you tell r what the sample is aes(sample=)
```{r}
chinstrap_penguins %>% 
  ggplot(aes(sample=body_mass_g))+
  geom_qq()+
  facet_wrap(~sex)
```

- as close to normally distributed as possible 



## Get Descriptive Statistics

```{r}
descriptive_stats<- chinstrap_penguins %>% 
  group_by(sex) %>% 
  summarise(
    mean_mass= mean(body_mass_g),
    median_mass= median(body_mass_g),
    sd_mass= sd(body_mass_g),
    sample_size= n())
```

- means are a bit different for females and males as we saw in our histograms

- both have sample size of 34

- similar standard deviation

- median is similar to mean- reflected in symmetric distribution in the histogram

## Get Vectors of body_mass_g for female and male chinstrap penguins seperatley: 

```{r}
chinstraps_f<- chinstrap_penguins %>% 
  filter(sex=="female") %>% 
  pull(body_mass_g)

chinstraps_m<- chinstrap_penguins %>% 
  filter(sex=="male") %>% 
  pull(body_mass_g)
```

## Run a two-sample tests, a couple different ways 

1. Is there a significant difference in body mass between male and female chinstrap penguins?

```{r}
chinstrap_ttest<- t.test(chinstraps_f, chinstraps_m)
chinstrap_ttest #call this to make visible
```

- mean of x (chinstraps_f)= 3527.206, mean of y (chinstraps_m= 3938.971

- null hypothesis: difference in means in not equal to 0, aka means are equal for male and females

- t= -5.2077 measure of how different from 0 our means are from the t-distribution

- df measure of sample size

- pvalue= very small probability that I could have gotten means at least this different by random chance if I drew samples from populations with the same mean --> sufficient evidence to reject null hypothesis --> conclude that there is a significant difference in body mass between male and female chinstraps 

- what is the actual difference? How much larger are male's than females? 

```{r}
cohen.d(chinstraps_f, chinstraps_m)
```

Cohen's d= -1.26 large effect size --> this is a clear difference that exists betweeen female and male chinstraps

There is a significant difference in body mass, and a large effect size (Cohen's *d* = -1.26). The way I'm *writing this* is not reproducible. How can we use in-line referencing of outcomes so that the values would automatically update if the values did? 

Use `broom::tidy()` to get messy model outputs into tidy data frames that are easy to reference! 

```{r}
chinstrap_out <- broom::tidy(chinstrap_ttest)
chinstrap_out
```

You can **reference values from model outputs in your R markdown text**:

t(`r round(chinstrap_out$parameter,2)`) = `r round(chinstrap_out$statistic,2)`, *p* = `r chinstrap_out$p.value`.

2. Are male chinstrap penguins significantly larger (in body mass) than female chinstrap penguins?

This is a *one-sided* test - we are only interested in the upper tail. We will still use the `t.test()` function, but update the default `alternative = ` argument from "two-sided" to "greater" or "less."

- one sided t-test
- order matters here 
- alternative= 'greater' or 'less' depending on which direction interested in

```{r}
# "Is mean male mass significantly GREATER than female mass?"
t.test(chinstraps_m, chinstraps_f, alternative = "greater")

# Yields the same statistical outcomes as "Is mean female mass significantly LESS than male mass?" 
t.test(chinstraps_f, chinstraps_m, alternative = "less")
```

- reject null that masses are equal, retain alt that male > female
- male chinstrap 


## One-way ANOVA

### 1. Make a table of counts by species and sex
- can use group_by(sex=="female), summarize(n())
- or can use count()

```{r}
penguin_counts <- penguins %>% 
count(species, sex)

penguin_counts
```

- large enough sample size that means are equally distributed 
- must also look at exploratory data viz to confirm 

Create a subset of only female penguins: 

Delineate species by: 
- aes(fill=species)
- +facet_wrap(~species)

- fill changes color of bars
- color changes color of points 
```{r}
penguin_f <-penguins %>% 
  filter(sex =="female")

#histogram
penguin_f %>% 
  ggplot(aes(x=flipper_length_mm))+ # don't give histogram 2nd variable
  geom_histogram(aes(fill=species))+ #colors by species
  facet_wrap(~species, scales="free")  # scales=free makes each panel different x and y scales
  
#qq-plot
penguin_f %>% 
  ggplot(aes(sample=flipper_length_mm))+ #feed qqplot a sample 
  geom_qq(aes(color=species))+ #colors by species
  facet_wrap(~species, scales="free")  # scales=free makes each panel different x and y scales
```

Closer these are to linear, closer to knowing that samples were drawn from a normally distributed population. 

### Find Summary Statistics

- standard error= sd/ sqrt(n)

When looking at equal variances, we will run a test for it. If your largest variance for a group is less than 4x your smallest variance for a group, then those are close enough. So in this case, it is within guidelines for close enough equality in variance for ANOVA test. 

### Levene's Test for Variance 
- as sample size gets larger, you will detect a significant difference in variance even if it is small or minimal
- make sure it's also within 4-to-1 guideline 

```{r}
penguin_f_summary<- penguin_f %>% 
  group_by(species) %>% 
  summarize(
    mean_flipper_length = mean(flipper_length_mm),
    sd_flipper_length = sd(flipper_length_mm),
    sample_size = n(),
    se_flipper_length = sd(flipper_length_mm) / sqrt(n()), 
    var_flipper_length = var(flipper_length_mm))

leveneTest(flipper_length_mm ~ species, data = penguin_f)
```

P-value (pr(>F)) is 0.056 is above significance level--> retain hypothesis that variance is equal

### Beeswarm plot with the mean +- 1sd

```{r}
ggplot()+ 
  geom_beeswarm(data=penguin_f, 
                aes(x=species, y=flipper_length_mm),
                color= "grey70",
                size= 3,
                alpha=0.5) +
  geom_point(data = penguin_f_summary, #adding the mean as a point on top of beeswarm
             aes(x=species, y=mean_flipper_length),
             color= "coral",
             size=5) +
  geom_errorbar(data=penguin_f_summary,
                aes(x=species,
                    ymax=mean_flipper_length + sd_flipper_length,
                    ymin=mean_flipper_length - sd_flipper_length),
                width=0.1)+ 
  theme_minimal()
```


## One-way ANOVA

Is there a significant different in mean flipper lengths in female penguins between Adelie, Chinstrap, and Gentoo penguins. 

- usually in in ENVS you want to do a type 3 ANOVA because sample sizes are unbalanced

- aov() is a type 1 ANOVA

```{r}
#run one way aov that compares flipper length as a function of species, only include female penguins data
penguin_aov <- aov(flipper_length_mm ~species, data=penguin_f)
summary(penguin_aov)

#post-hoc test: Tukey's HSD
#does pairwise comparisons that adjusts so no .05 error 
TukeyHSD(penguin_aov)
#returns adjusted p-values 
#there is a sig difference between all of these groups 
```

F-statistic is the ratio of between group sum of squares divided within group sum of squares. In this case, this is the difference between groups is very large than within groups. This is reflected in the p-value, very low. Negligably low probability that we could gotten means this different if they were actually were from populations with the same mean. Reject null hypothesis that they have equal means. 

Post hoc test returns adjusted p-values.
Shows that there is a significant difference between all of the group's flipper length.

## Type 3 ANOVA

If you have very different sample sizes and distributions. Orthogonial or type 3 anova. 
When doing this in the real world, look up which to use 

```{r}
penguin_lm<- lm(flipper_length_mm ~species, data = penguin_f)  #linear model
penguin_aov_type3<- Anova(penguin_lm, type = "III")
penguin_aov_type3

```

The F value is the same as the type 1 we ran above. Should be very similar unless data has unbalanced design and very different sample sizes within groups. 