---
title: "ESM 206 Week 5 Coding Lab"
author: "Shelby Smith"
date: "11/10/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
library(effsize)
library(janitor)
library(broom)
library(ggbeeswarm)
library(car)
```

# Part 1. T-tests 

Mass comparison between male and female chinstrap penguins

```{r}
chinstrap_penguins <- penguins %>% 
  filter(species=="Chinstrap")

# Histograms (only takes one variable)
chinstrap_penguins %>% 
  ggplot(aes(x= body_mass_g)) +
  geom_histogram(bins=10)+
  facet_wrap(~sex)
  
```

- distribution overall looks pretty normal generally bell shaped, so mean would be a reasonable measure of central tendency

- male central tendency is higher (about 4000), and female central tendency is lower around 3500

## QQ plot

- #in qqplot, you tell r what the sample is aes(sample=)
```{r}
chinstrap_penguins %>% 
  ggplot(aes(sample=body_mass_g))+
  geom_qq()+
  facet_wrap(~sex)
```

- as close to normally distributed as possible 



## Get Descriptive Statistics

```{r}
descriptive_stats<- chinstrap_penguins %>% 
  group_by(sex) %>% 
  summarise(
    mean_mass= mean(body_mass_g),
    median_mass= median(body_mass_g),
    sd_mass= sd(body_mass_g),
    sample_size= n())
```

- means are a bit different for females and males as we saw in our histograms

- both have sample size of 34

- similar standard deviation

- median is similar to mean- reflected in symmetric distribution in the histogram

## Get Vectors of body_mass_g for female and male chinstrap penguins seperatley: 

```{r}
chinstraps_f<- chinstrap_penguins %>% 
  filter(sex=="female") %>% 
  pull(body_mass_g)

chinstraps_m<- chinstrap_penguins %>% 
  filter(sex=="male") %>% 
  pull(body_mass_g)
```

## Run a two-sample tests, a couple different ways 

1. Is there a significant difference in body mass between male and female chinstrap penguins?

```{r}
chinstrap_ttest<- t.test(chinstraps_f, chinstraps_m)
chinstrap_ttest #call this to make visible
```

- mean of x (chinstraps_f)= 3527.206, mean of y (chinstraps_m= 3938.971

- null hypothesis: difference in means in not equal to 0, aka means are equal for male and females

- t= -5.2077 measure of how different from 0 our means are from the t-distribution

- df measure of sample size

- pvalue= very small probability that I could have gotten means at least this different by random chance if I drew samples from populations with the same mean --> sufficient evidence to reject null hypothesis --> conclude that there is a significant difference in body mass between male and female chinstraps 

- what is the actual difference? How much larger are male's than females? 

```{r}
cohen.d(chinstraps_f, chinstraps_m)
```

Cohen's d= -1.26 large effect size --> this is a clear difference that exists betweeen female and male chinstraps

There is a significant difference in body mass, and a large effect size (Cohen's *d* = -1.26). The way I'm *writing this* is not reproducible. How can we use in-line referencing of outcomes so that the values would automatically update if the values did? 

Use `broom::tidy()` to get messy model outputs into tidy data frames that are easy to reference! 

```{r}
chinstrap_out <- broom::tidy(chinstrap_ttest)
chinstrap_out
```

You can **reference values from model outputs in your R markdown text**:

t(`r round(chinstrap_out$parameter,2)`) = `r round(chinstrap_out$statistic,2)`, *p* = `r chinstrap_out$p.value`.

2. Are male chinstrap penguins significantly larger (in body mass) than female chinstrap penguins?

This is a *one-sided* test - we are only interested in the upper tail. We will still use the `t.test()` function, but update the default `alternative = ` argument from "two-sided" to "greater" or "less."

- one sided t-test
- order matters here 
- alternative= 'greater' or 'less' depending on which direction interested in

```{r}
# "Is mean male mass significantly GREATER than female mass?"
t.test(chinstraps_m, chinstraps_f, alternative = "greater")

# Yields the same statistical outcomes as "Is mean female mass significantly LESS than male mass?" 
t.test(chinstraps_f, chinstraps_m, alternative = "less")
```
```

- reject null that masses are equal, retain alt that male > female
- male chinstrap 