---
title: "Week 5 Coding Lesson - T-tests, Cohen's d, one-way ANOVA"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(palmerpenguins)
library(effsize)
library(broom)
library(ggbeeswarm)
library(car)

```

## Part 1. T-tests & effect size


Mass difference: male & female chinstrap penguins

```{r}
chinstraps <- penguins %>% 
  filter(species == "Chinstrap")

# Look at it:

# Histograms
ggplot(data = chinstraps, aes(x = body_mass_g)) +
  geom_histogram(bins = 12) +
  facet_wrap(~sex)

# QQ Plots
ggplot(data= chinstraps, aes(sample = body_mass_g)) +
  geom_qq() +
  facet_wrap(~sex)

# Get some summary statistics:
chinstraps %>% 
  group_by(sex) %>% 
  summarize(mean_mass = mean(body_mass_g),
            median_mass = median(body_mass_g),
            sd_mass = sd(body_mass_g),
            sample_size = n())
```

### Get the vectors for body mass values: 

```{r}
chinstraps_f <- chinstraps %>% 
  filter(sex == "female") %>% 
  pull(body_mass_g)

chinstraps_m <- chinstraps %>% 
  filter(sex == "male") %>% 
  pull(body_mass_g)
```

### T-tests, a couple different ways: 

1. Is there a significant difference in body mass between male and female chinstrap penguins?
```{r}
chinstrap_ttest <- t.test(chinstraps_f, chinstraps_m)
chinstrap_ttest

cohen.d(chinstraps_f, chinstraps_m)
```

There is a significant difference in body mass, and a large effect size (Cohen's *d* = -1.26). The way I'm *writing this* is not reproducible. How can we use in-line referencing of outcomes so that the values would automatically update if the values did? 

Use `broom::tidy()` to get messy model outputs into tidy data frames that are easy to reference! 

```{r}
chinstrap_out <- broom::tidy(chinstrap_ttest)
chinstrap_out
```

You can **reference values from model outputs in your R markdown text**:

t(`r round(chinstrap_out$parameter,2)`) = `r round(chinstrap_out$statistic,2)`, *p* = `r chinstrap_out$p.value`.

2. Are male chinstrap penguins significantly larger than female chinstrap penguins? 

This is a *one-sided* test - we are only interested in the upper tail. We will still use the `t.test()` function, but update the default `alternative = ` argument from "two-sided" to "greater" or "less."

```{r}
# "Is mean male mass significantly GREATER than female mass?"
t.test(chinstraps_m, chinstraps_f, alternative = "greater")

# Yields the same statistical outcomes as "Is mean female mass significantly LESS than male mass?" 
t.test(chinstraps_f, chinstraps_m, alternative = "less")
```

## One-way ANOVA

### 1. Make a table of counts by species and sex

```{r}
penguin_counts <- penguins %>% 
  count(species, sex)

penguin_counts
```

We ask: Is there a significant difference in mean flipper length (mm) between the three species for **female** penguins?

- Null hypothesis: There is no significant difference in mean flipper length between the three groups

- Alternative hypothesis: There **is** a significant difference in mean flipper length between the three groups

### 2. First, look at the distributions for female penguin flipper lengths

```{r}
penguin_f <- penguins %>% 
  filter(sex == "female")

# Faceted histograms: 
ggplot(data = penguin_f, aes(x = flipper_length_mm)) +
  geom_histogram(aes(fill = species),
                 alpha = 0.5,
                 show.legend = FALSE,
                 bins = 15) +
  facet_wrap(~species, scales = "free")

# Faceted QQ plots: 
ggplot(data = penguin_f, aes(sample = flipper_length_mm)) +
  geom_qq(aes(color = species),
                 alpha = 0.5,
                 show.legend = FALSE
          ) +
  facet_wrap(~species, scales = "free")

# These look pretty darn normal. Let's see what their means, sds, se, and variances are: 

```

### 3. Find summary statistics, check for equal variances

```{r}
penguin_f_summary <- penguin_f %>% 
  group_by(species) %>% 
  summarize(
    mean_flipper_length = mean(flipper_length_mm),
    sd_flipper_length = sd(flipper_length_mm),
    sample_size = n(),
    se_flipper_length = sd(flipper_length_mm) / sqrt(n()),
    var_flipper_length = var(flipper_length_mm)
  )

penguin_f_summary

# Importantly, variances are pretty close (largest is < 4x greater than the smallest)
# But we can perform a Levene's Test for equal variances anyway: 

leveneTest(flipper_length_mm ~ species, data = penguin_f) # Not significant (we retain the null hypothesis of equal variances)

```

### 4. Make a plot of data with mean and SE indicated: 

```{r}

ggplot() +
  geom_beeswarm(data = penguin_f, 
                aes(x = species, 
                    y = flipper_length_mm,
                    ),
                size = 3,
                alpha = 0.6,
                pch = 16,
                color = "gray70") +
  scale_x_discrete(labels = c("Adelie", "Chinstrap", "Gentoo")) +
  geom_point(data = penguin_f_summary, aes(x = species, y = mean_flipper_length), 
             color = "firebrick",
             size = 4) +
  geom_errorbar(data = penguin_f_summary, 
                aes(x = species, 
                    ymin = mean_flipper_length - sd_flipper_length,
                    ymax = mean_flipper_length + sd_flipper_length),
                color = "firebrick",
                width = 0.1) +
  theme_light()

```

### 5. One-way ANOVA

Use `aov()` for one-way (single factor) ANOVA in R:

```{r}

penguin_aov <- aov(flipper_length_mm ~ species, data = penguin_f)
summary(penguin_aov)

# We retain the alternative hypothesis (mean flipper lengths for the 3 species are NOT the same). But are they all different from each other? Or just two? 

```

##### Additional: comparing to Type III ANOVA

Note: for unbalanced groups, you may want to do Type III ANOVA instead of Type I (they will yield the same results for balanced design, but for very unbalanced groups you probably want Type III).

```{r}

penguin_lm <- lm(flipper_length_mm ~ species, data = penguin_f)
penguin_aov_t3 <- Anova(penguin_lm)
penguin_aov_t3

# Notice that the F statistic value here (411.79) is almost exactly the same - the difference between these two is minimal. 
```


### 6. Post-hoc Tukey's HSD

```{r}
TukeyHSD(penguin_aov)

# Conclusion: means for ALL pairwise combinations are significantly different
```

Conclusion: Mean flipper lengths differed significantly between Chinstrap (mean +/- se), Gentoo (mean +/- se), and Adelie (mean +/- se) penguins (one-way ANOVA with post-hoc Tukey's HSD, F(df) = ??, *p* < 0.05). 

### END
